---
title: "Final Project - Predicting March Madness"
author: "Anmol Sapru and Rohit Gunda"
format: pdf
---

# [Introduction and Data]{.underline}

Duke is synonymous with basketball. As Duke students who love Duke Basketball and March Madness, we are interested in performing a statistical analysis on the most thrilling tournament in sports. While watching the 2023 March Madness tournament and the many upsets that came with it, we were motivated to see if we could use statistical methods to predict March Madness winners. Upon scouring sites such as FiveThirtyEight and the KenPom rankings, we were inspired to create models of our own to predict tournament success.

**Fundamental Research Question:** What variables are important to March Madness success and which outliers over the past 8 years have existed that bring the "Madness" to the tournament?

The following packages were used:

```{r load-packages, message = FALSE, warning = FALSE}

library(tidyverse)
library(tidymodels)
library(Stat2Data)
library(caret)
library(leaps)
library(MASS)
```

Data was found from the [Sports Reference college basketball team stats website](https://www.sports-reference.com/cbb/seasons/men/2023-school-stats.html) and [Bart Torvik analytics website](https://barttorvik.com/#), with these general and deeper stats being taken going back to 2008 (excluding the cancelled March Madness of 2020). From here, the data was cleaned in Excel to separate the year and result of each team then joined for all of this data together.

Key Variables

-   

**EXPLORATORY DATA ANALYSIS**

For further cleaning, the non-March Madness teams added in the join were removed and simple variable names and values were cleaned up to be easier to work with.

**ADD CLEARER DESCRIPTION OF DATA CLEANING**

```{r load-data, echo = FALSE, message = FALSE}

cbb <- read_csv("data/20082022torvik.csv") 
sportsreference <- read_csv("data/sportsreference.csv")
background <- read_csv("data/teams_background.csv")
sportsreference$Year <- sportsreference$Year-1

cbb <- left_join(cbb, background, by = c("TEAM" = "TEAM", `...23` = "YEAR"))

#Remove non-postseason teams and R68 losers
cbb <- cbb[!is.na(cbb$march_madness),]
cbb <- filter(cbb, !grepl("R68", march_madness))

#Cleaning up variable names, variables, etc
cbb$march_madness <- str_trim(cbb$march_madness, side = c("both"))
cbb <- rename(cbb, march_madness = march_madness) 
cbb <- rename(cbb, "YEAR" = `...23`) 

cbb <- cbb |>
  mutate(march_madness = case_when(
    march_madness == "Sweet Sixteen" ~ "S16",
    march_madness == "Elite Eight" ~ "E8",
    march_madness == "Final Four" ~ "F4",
    march_madness == "Finals" ~ "2ND",
    march_madness == "CHAMPS" ~ "Champions",
    TRUE ~ march_madness
  ))

cbb <- left_join(cbb, sportsreference, by = c("TEAM" = "School", "YEAR" = "Year"))
```

# [Methodology]{.underline}

We determined it would be best to try different regressions, with logistic regressions by round and an ordinal regression---each with their different assumptions to look into. The issue with the first of these is the low levels of data that limit the creation of a model beyond around the Elite Eight. In preparing for the round-by-round logistic regressions we split up the data from the original data sets into multiple data sets for each round, with a TRUE or FALSE value of whether they won their game in that round.

```{r round-separation, echo = FALSE}

#separated each row by round for determining differences
cbb_r <- mutate(cbb, round_64 = if_else(march_madness == "R64", FALSE, TRUE))
round_64 <- cbb_r

cbb_r <- mutate(cbb_r, round_32 = 
         case_when(march_madness == "R32" ~ FALSE,
         march_madness %in% c("S16", "E8", "F4", "2ND", "Champions") ~ TRUE,
         TRUE ~ NA))
round_32 <- cbb_r[!is.na(cbb_r$round_32),]

cbb_r <- mutate(cbb_r, sweet_sixteen = 
         case_when(march_madness == "S16" ~ FALSE,
         march_madness %in% c("E8", "F4", "2ND", "Champions") ~ TRUE,
         TRUE ~ NA))
sweet_sixteen <- cbb_r[!is.na(cbb_r$sweet_sixteen),]

cbb_r <- mutate(cbb_r, elite_eight = 
         case_when(march_madness == "E8" ~ FALSE,
         march_madness %in% c("F4", "2ND", "Champions") ~ TRUE,
         TRUE ~ NA))
elite_eight <- cbb_r[!is.na(cbb_r$elite_eight),]

cbb_r <- mutate(cbb_r, final_four = 
         case_when(march_madness == "F4" ~ FALSE,
         march_madness %in% c("2ND", "Champions") ~ TRUE,
         TRUE ~ NA))
final_four <- cbb_r[!is.na(cbb_r$final_four),]

cbb_r <- mutate(cbb_r, champ_game = 
         case_when(march_madness == "2ND" ~ FALSE,
         march_madness %in% c("Champions") ~ TRUE,
         TRUE ~ NA))
champ_game <- cbb_r[!is.na(cbb_r$champ_game),]
```

**WHY CHOSE EACH VARIABLE**

## Round-by-Round Logistic Regression

The following is an example of the regression that was ran on all of the Round of 64 teams to create a regression that predicts winners (round_64 = TRUE) against losers (round_64 = FALSE) for the round. Using stepAIC works to attempt to limit the overfitting with the data. This was also done for the Round of 32 and Sweet Sixteen.

```{r round-64-model, results = "hide"}

round_64 <- na.omit(round_64)

round_64_max <- glm(round_64 ~ G.x + WINS + LOSSES + ADJOE + ADJDE + 
                      `EFG%` + `EFGD%` + TOR + TORD + ORB + DRB + FTR +
                      FTRD + `2P%` + `2P%D` + `3P%` + `3P%D` + `3PR` +
                      `3PRD` + `ADJ T.` + `Overall SRS` + `Overall SOS` +   
                      `Conf. W-L%` + `Home W-L%` + `Away W-L%` + 
                      `AVG PPG` + `AVG DPPG` + `AVG PD` + `AST/TOV` + 
                      `PF/G` + WINS*G.x + LOSSES*G.x + ADJOE*ADJDE +
                      `EFG%`*`EFGD%` + TOR*TORD + ORB*DRB + FTR*FTRD +
                      `2P%`*`2P%D` + `3P%`*`3P%D` + `3PR`*`3PRD` +
                      `2P%`*`3P%` + `2P%D`*`3P%D` + `AVG PPG`*`AVG DPPG` +
                      WINS*`Overall SRS` + WINS*`Overall SOS` +
                      LOSSES*`Overall SRS` + LOSSES*`Overall SOS`,     
                      data = round_64,
                      family = "binomial")

round_64_min <- glm(round_64 ~ 1,
                    data = round_64,
                    family = "binomial")

round_64_model <- stepAIC(round_64_max,
        scope = list(lower = round_64_min, upper = round_64_max),
        data = round_64, direction = "both")
```

```{r round-32-model, echo = FALSE, results = "hide"}

round_32_max <- glm(round_32 ~ ADJOE + ADJDE + `EFG%` + `EFGD%` +
                      TOR + TORD + ORB + DRB + FTR + FTRD + `2P%` + 
                      `2P%D` + `3P%` + `3P%D` + `ADJ T.` + 
                      `Overall SRS` + `Overall SOS` + `Conf. W-L%` + 
                      `Home W-L%` + `Away W-L%` + `AVG PPG` + 
                      `AVG DPPG` + `AVG PD` + `AST/TOV` + `PF/G` + 
                      ADJOE*ADJDE + `EFG%`*`EFGD%` + TOR*TORD + 
                      ORB*DRB + FTR*FTRD + `2P%`*`2P%D` + 
                      `2P%`*`3P%` + `3P%`*`3P%D` + `AVG PPG`*`AVG DPPG`,
                      data = round_32,
                      family = "binomial")

round_32_min <- glm(round_32 ~ 1,
                    data = round_32,
                    family = "binomial")

round_32_model <- stepAIC(round_32_max,
        scope = list(lower = round_32_min, 
                     upper = round_32_max),
        data = round_32, direction = "both")
```

```{r sweet-sixteen-model, echo = FALSE, results = "hide"}

sweet_sixteen_max <- glm(sweet_sixteen ~ ADJOE + ADJDE + `EFG%` + `EFGD%` +
                      TOR + TORD + ORB + DRB + FTR + FTRD + `2P%` + 
                      `2P%D` + `3P%` + `3P%D` + `ADJ T.` + 
                      `Overall SRS` + `Overall SOS` + `Conf. W-L%` + 
                      `Home W-L%` + `Away W-L%` + `AVG PPG` + 
                      `AVG DPPG` + `AVG PD` + `AST/TOV` + `PF/G` + 
                      ADJOE*ADJDE + `EFG%`*`EFGD%` + TOR*TORD + 
                      ORB*DRB + FTR*FTRD + `2P%`*`2P%D` + 
                      `2P%`*`3P%` + `3P%`*`3P%D` + `AVG PPG`*`AVG DPPG`,
                        data = sweet_sixteen,
                        family = "binomial")

sweet_sixteen_min <- glm(sweet_sixteen ~ 1,
                    data = sweet_sixteen,
                    family = "binomial")

sweet_sixteen_model <- stepAIC(sweet_sixteen_max,
        scope = list(lower = sweet_sixteen_min,
                     upper = sweet_sixteen_max),
        data = sweet_sixteen, direction = "both")
```

The following are the regression models that were found from the three rounds of data and regression models run through stepAIC:

## Overall Ordinal Regression

```{r ordinal-reg}

cbb$march_madness <- fct_relevel(cbb$march_madness, 
                                 c("R64", "R32", "S16", "E8", 
                                   "F4", "2ND", "Champions"))

mmb_ord_max <- polr(march_madness ~ 
                     G.x + WINS + LOSSES + ADJOE + ADJDE + 
                      `EFG%` + `EFGD%` + TOR + TORD + ORB + DRB + FTR +
                      FTRD + `2P%` + `2P%D` + `3P%` + `3P%D` + `3PR` +
                      `3PRD` + `ADJ T.` + `Overall SRS` + `Overall SOS` +   
                      `Conf. W-L%` + `Home W-L%` + `Away W-L%` + 
                      `AVG PPG` + `AVG DPPG` + `AVG PD` + `AST/TOV` + 
                      `PF/G` + WINS*G.x + LOSSES*G.x + ADJOE*ADJDE +
                      `EFG%`*`EFGD%` + TOR*TORD + ORB*DRB + FTR*FTRD +
                      `2P%`*`2P%D` + `3P%`*`3P%D` + `3PR`*`3PRD` +
                      `2P%`*`3P%` + `2P%D`*`3P%D` + `AVG PPG`*`AVG DPPG` +
                      WINS*`Overall SRS` + WINS*`Overall SOS` +
                      LOSSES*`Overall SRS` + LOSSES*`Overall SOS`, 
                   data = cbb)

mmb_ord_min <- polr(march_madness ~ 1, data = cbb)

mmb_ord_model <- stepAIC(mmb_ord_max,
        scope = list(lower = mmb_ord_min,
                     upper = mmb_ord_max),
        data = cbb, direction = "both")
```

**ADD ASSESSMENT OF CONDITIONS + DIAGNOSTICS**

# [Results]{.underline}

```{r predictions}

`2023sr` <- read_csv("data/2023sportsreference.csv")
`2023analytics` <- read_csv("data/2023torvik.csv")

`2023stats` <- left_join(`2023analytics`, `2023sr`, by = c("TEAM" = "School"))

`2023teams` <- read_csv("data/2023teams.csv")

`2023stats` <- `2023stats` |>
  filter(`2023stats`$TEAM %in% `2023teams`$TEAM)

tibble(predict(round_64_model, `2023stats`)) |>
  mutate(rank = seq(1:64)) |>
  left_join(mutate(`2023stats`, rank = seq(1:64))) |>
  arrange(desc(predict(round_64_model, `2023stats`)))

tibble(predict(round_32_model, `2023stats`)) |>
  mutate(rank = seq(1:64)) |>
  left_join(mutate(`2023stats`, rank = seq(1:64))) |>
  arrange(desc(predict(round_32_model, `2023stats`)))

tibble(predict(sweet_sixteen_model, `2023stats`)) |>
  mutate(rank = seq(1:64)) |>
  left_join(mutate(`2023stats`, rank = seq(1:64))) |>
  arrange(desc(predict(sweet_sixteen_model, `2023stats`)))

tibble(predict(mmb_ord_model, `2023stats`)) |>
  mutate(rank = seq(1:64)) |>
  left_join(mutate(`2023stats`, rank = seq(1:64))) |>
  arrange(desc(predict(mmb_ord_model, `2023stats`, type = "probs")))
```

**INTERPRET SOMETHING - MOST IMPORTANT VARIABLES**

**ADD ASSUMPTIONS**

**MODEL'S PREDICTIVE POWER**

## Outliers and Randomness of March Madness

# [Discussion]{.underline}

**MISSING ASSUMPTIONS**

**HYPOTHESIS TESTS**

**COOK'S DISTANCE**

**IDEA'S FOR FUTURE WORK**
